Outline flowCore paper

### 2008-04-30 paper/manuscript.tex
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/flowCore@31457 bc3139a8-67e5-0310-9ffc-ced21a209358

\section*{Introduction}
Traditionally, flow cytometry (FCM) has been a tube-based technique
limited to small-scale laboratory and clinical studies.  High
throughput methods for flow cytometry (FC-HCS) have recently been
developed and are now used in both basic and clinical research. Today,
the need for flexible and well structured computational tools to
efficiently handle and analyze FC-HCS has increased
considerably. Indeed, the amount of information generated by these
technologies must be stored, managed, and needs to be summarized in
order to be accessible to the researcher. However, the absence of a
shared research platform that bioinformaticians, computer scientists,
and statisticians can utilize to develop novel or standard methods for
flow cytometric analysis has hindered innovation. We believe that the
open source statistical software R in conjunction with the
Bioconductor Project can fill this gap.  In this paper, we propose R
data structures to handle flow cytometry data through the main steps
of importing, storing, assessing and preprocessing. These structures,
along with a range of specialized methods for compensation,
transformation, gating and additional preprocessing steps, are
implemented in the newly developed Bioconductor package flowCore.  We
envision flowCore to be a solid foundation for researchers to
efficiently handle FC-HCS data and for future development of new
methods and tools for their coherent analysis.

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In addition, R is an open source research
platform which bioinformaticians, computer scientists, and
statisticians routinely use to develop novel tools and methodology for
statistical data analysis. For biological applications, the
Bioconductor project has become the standard toolset
\citep{gentleman2006bos} to process data spanning a diversity of
research fields from genomics to proteomics to complex cell
biology. It is particularly geared towards the analysis of large
numeric datasets that typically arise from high throughput experiments
such as microarray gene expression analysis or FC-HCS. A shared
platform proved to be particularly beneficial for the development of
analysis routines for microarray gene expression data, and by today
many of these early innovations have been incorporated into commercial
software products, thus greatly improving the quality of gene
expression data analysis. Shared data structures have been shown to be
essential for such a collaborative effort. We envision a similar
paradigm for FC-HCS data.

\subsection*{Existing data standards and conventions}
Currently, data from flow cytometry experiments are stored in single
files following a well-defined data standard (FCS). This standard
attempts to capture much of the accompanying meta information, most of
which is produced by the measurement instrument or has to be supplied
manually by the operator upon experiment setup. However, the recent
developments in high-throughput flow cytometry are shifting the focus
of interest away from single-tube based measurements towards large and
complex experimental designs with dozens of co-variates and
influencing factors. Self-contained and self-reflective data
structures are a prerequisite to allow for coordinated analysis of
such experiments. Most of the currently available software solutions
offer only limited support for such structures, or make use of
containers like XML or binary storage formats that are designed
specifically for the mostly GUI-driven user interfaces and hence are
not easily amenable to programmatical access. In addition, the
closed-source nature of most of these products makes it impratical to
connect to and integrated the data into an analysis pipeline.

Traditionally, the majority of flow cytometry experiments have been
analyzed manually by direct data inspection in one or two dimensions,
or by very basic comparisons of summary statistics. We believe that
these approaches do not fully address the highly complex nature of FCM
data, in particular, they discard many of the fundamental aspects of
the data, such as its underlying distribution and its high-dimensional
nature. In addition, the subjective character of manual analyses are a
major obstacle to reproducibility. For FC-HCS data, unassisted manual
inspection has become way too time and labor intensive, and robust
statistical methods need to be developed to point the investigator to
the interesting aspects of the data or to potential problems. While
the expert knowledge of immunologists and researchers remains crucial
for the understanding of FCM data, we believe that tight collaboration
with other research fields such as statistics and computer science can
greatly improve the relevance of flow cytometry in today's
high-throughput paradigm.

These issues are currently being addressed by emerging new flow
cytometry data standards developed in collaboration with the ISAC Data
Standards Task Force.  The Gating-ML Candidate Recommendation (CR)
represents a proposal on how to form unambiguous XML-based gate
definitions, which can facilitate the interchange and validation of
data between different software packages with the potential of
significant increase of hardware and software interoperability. Gates
may be ordered into a hierarchical structure to describe a gating
strategy. They may be applied on parameters as in FCS files or on
transformed parameters as described by an explicit parameter
transformation. This enables the encoding of a simple analytical work
flow, in a way that allows us to reconstruct the analysis
programmatically.

The flowCore framework presented here offers import functionality for
raw data FCS files along with their complete set of file-specific meta
information (Figure~\ref{fig1:FrameWork}). Moreover, it is a software
implementations of the Gating-ML CR, which opens the possibility to
integrate flowCore in existing work flows and to communicate with
virtually every other FCM tool that adheres to the proposed
standard. In addition, the software is platform agnostic, running on
Windows, Mac OS X and Linux/Unix operating systems.

flowCore is not a GUI-driven software, and all operations are done
using a command line interface. Conceptually, it is possible to add a
more elaborate user interface on top of that, however the focus is on
a programmatic approach to enable the convenient development of novel
analysis methods. By taking the burden of data management from the
programmer and by providing well-defined APIs and a structured class
hierarchy, it is possible to readily test new ideas and to easily
extend the frameworks functionality. Both the underlying R and
Bioconductor software and flowCore itself are open source and open
development projects, thus community involvement in the developing
process is highly appreciated.

\section*{Basic data structures}
\subsection*{Flow Frame}
flowCore's primary task is the representation and basic manipulation
of flow cytometry data. This is accomplished through a data model very
close to that adopted by other Bioconductor packages used for gene
expression analysis, and thus familiar to most Bioconductor
users. Similar to any other flow tool, the basic unit of organization
in flowCore is a collection of events. We call the structure that hold
this data a flow frame. All relevant information for the collection
are stored along with the matrix of raw data values and can be
accessed programmatically. Most commonly, these are descriptors of the
stains used in the experiment and the respective measurement channels,
information about compensation performed at the instrument side and
any additional keywords the user deems to be important to annotate the
data. A number of quality checks are performed during the creation of
a flow frame to ensure data integrity. There is an extended interface
that allows for the standardized interaction with the flow data, thus
enabling the development of more specialized tools for specific tasks
of the data analysis process that all share a common API.

\subsection*{Flow Set}
In high-throughput flow cytometry, many of the analysis tasks need to
be performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flow frames called a flow
set. Flow sets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flow frames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of mixups and errors that are
all too common when using more or less ad hoc spreadsheets that lack a
controlled structure. In general, there is no native ordering to a
flow set, which in a way reflects the still predominant application of
tube-based experiments rather than experiments following a
plate-design. However, the flow set structure can be readily extended
to also accommodate such spatial information \textit{(plateCore
  reference, Perry, Errol?)}.

%%Figure1 Diagram of the data structures and the basic operations
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
\caption{\label{fig1:FrameWork}{The flowCore frame work.For each
    experiement, FCS files, phenotypic and meta data is store in a
    flow set. Each flow frame of a flow set corresponds to one FCS
    file. All basic operation (compensation, transformation, gating,
    ...) can be applied on both single flow frames or entire flow
    set.}}
\end{figure}

\section*{Standard flow operations}
The basic operations in flow cytometry analysis are typically the
same: the data need to be compensated (if not already done on the
instrument side), often transformed and sup-population of interest
need to be selected based on a set of (predominantly sequential)
gates. All software solutions for flow data analysis offer more or
less elaborate support for these kinds of operation, most of them in
an interactive, GUI-driven spirit. The approach we have taken in
flowCore is to abstractly describe these operations and build a set of
tools to perform them on both flow frames as well as on entire flow
sets. While transformation and to a certain extent also compensation
are fairly routine operations with only limited potential for
improvement, being able to implement new methodology for gating of
flow cytometry data, and extending the capabilities of flowCore
through object oriented programming are features that clearly sets our
framework apart from other flow toolsets around. By factoring out as
much of the bookkeeping as possible, programmers can focus on the
actual operations rather then having to deal with the tedious details
of data integration and access. Third party methods implemented by
others can act on their own as first-class citizens in the analysis
framework, without breaking the workflow or the basic
infrastructure. This design allows for the straightforward extension
of flowCore's capabilities and already fostered the development of a
number of valuable improvements. \citep{lo2008agf,sarkar2008ufv}
\textit{(plateCore reference, Perry, Errol?)}.

\subsection*{Transformation and compensation}
Proper transformation have been proven beneficial for both data
visualization and modelling \citep{lo2008agf}. The major
transformations that are routinely used in flow cytometry analysis
have been implemented in flowCore. Furthermore, the design of the R
language makes it easy to define arbitrary functions to apply to the
data of individual flow frames or entire flow sets,
respectively. Compensation is available for both flow frames and flow
sets, and the software also offers functionality to compute spillover
or compensation matrices from a set of appropriate compensation
samples.

@@

Static gating for all samples in a high-throughput flow cytometry
experiment is often not an option, since the measured variables tend
to vary between different treatments or over time. Automated or
data-driven gating tries to estimate the gating regions from the
underlying data, thus providing a fast objective solution to the
analysis of potentially very large and diverse data sets
\cite{lo2008agf}. One of the automated gating methods implemented in
flowCore is based on identifying areas of significant curvature in a
kernel density estimate of the data \citep{wand2008}. Assuming that
the regions of interest are of high density, the software is able to
reliably detect them in either the one ore the two-dimensional density
landscape. Kernel density estimation is a well-known problem in
statistical computing, and a lot of effort has been invested in the
development of good software to address it. The modular design of R,
Bioconductor and flowCore allows to easily integrate these existing
solutions into our framework. In this example, we directly use R code
from the feature package developed by Wand et al. Instead of
re-writing existing code, we are able to include it via the well
tested distribution mechanism provided by R's software package
system. This process is bi-directional, and all funtionality
implemented in flowCore is available to other package authors, as we
have seen with the afore-mentioned flowClust package.

In developing flowCore, we tried to provided a unified develoment and
analysis platform for flow-cytometry data. The scope and complexity of
the technology is evolving rapidly, and a collaborative effort in
devising new methodolgy has been proven beneficial for a number of
different biological and computational biology challenges. We hope
that our framework will be the foundation for fruitful shared research
by many calloborators from multiple scientific fields, further pushing
the boundaries of the exiting and still promising flow cytometry
technology.

>>>>

\section*{Introduction}
Traditionally, flow cytometry (FCM) has been a tube-based technique
limited to small-scale laboratory and clinical studies.  High
throughput methods for flow cytometry high content screening (FC-HCS) have recently been
developed and are now used in both basic and clinical research. As a result, the need for flexible and well structured computational tools to
efficiently handle and analyze FC-HCS has increased
considerably. 
%Indeed, the amount of information generated by these technologies must be stored, managed, and needs to be summarized in order to be accessible to the researcher. This sentance is not grammaically corect, and really doesn't add anything anyway
However, the absence of a
shared research platform that bioinformaticians, computer scientists,
and statisticians can utilize to develop novel %or standard - not sure we want to advocate standard methods? Do we really want to say everyone should use the same method for analysis?
methods for
flow cytometric analysis has hindered innovation. <ref Lizard G. FCM analyses and bioinformatics: Interest in new softwares to
optimize novel technologies and to favor the emergence of innovative concepts in cell
research. Cytometry A 2007;71A:646–647>
We believe that the
open source statistical software R, in conjunction with the
Bioconductor Project can fill this gap.  In this paper, we propose R
data structures to handle FCM data through the main steps
of importing, storing, assessing and preprocessing. These structures,
along with a range of specialized methods for compensation,
transformation, gating and additional preprocessing steps, are
implemented in the newly developed Bioconductor package flowCore.  flowCore is a freely available, sophisticated and extensible FCM data analysis platform that enables researchers to
efficiently handle FC-HCS data and enables open development of new
methods and tools for their coherent analysis. flowCore runs on
Windows, Mac OS X and Linux/Unix operating systems.

\subsection*{R and Bioconductor}
R is a robust statistical programing environment which offers a wide
range of statistical and visualization methods developed for various
fields of application. In addition, R is an open source research %already mentioned R, and by extention flowCore open source
platform that bioinformaticians, computer scientists, and
statisticians routinely use to develop novel tools and methodology for
statistical data analysis. For biological applications, the
Bioconductor project has become the standard toolset
\citep{gentleman2006bos} to process data spanning a diversity of
research fields from genomics to proteomics to complex cell
biology. It is particularly geared towards the analysis of large
numeric datasets that typically arise from high throughput experiments
such as microarray gene expression analysis or FC-HCS. A shared
platform proved to be particularly beneficial for the development of
analysis routines for microarray gene expression data, and by today
many of these early innovations have been incorporated into commercial
software products, thus greatly improving the quality of gene
expression data analysis. Shared data structures have been shown to be
essential for such a collaborative effort. We envision a similar
paradigm for FC-HCS data.

\subsection*{Existing data standards and conventions}
Currently, data from FCM experiments are stored in single
files according to the Flow Cytometry Standard (FCS)<add ref to FCS3 Cytometry article>. 
%This standard attempts to capture much of the accompanying meta information, most of which is produced by the measurement instrument or has to be supplied manually by the operator upon experiment setup. - I believe FCS only tries to capture a subset of meta information
However, the recent
developments in high-throughput FCM are shifting the focus
of interest away from single-tube based measurements towards large and
complex experimental designs with dozens of co-variates and
influencing factors. Self-contained and self-reflective data
structures are a prerequisite to allow for coordinated analysis of
such experiments. Most of the currently available software solutions
offer only limited support for such structures, or make use of
containers like XML or binary storage formats that are designed
specifically for the mostly GUI-driven user interfaces and hence are
not easily amenable to programmatical access. In addition, the
closed-source nature of %most of <can you think of an example that is not closed source?>
these products makes it impractical to
connect to and integrate the data into an analysis pipeline.

Traditionally, the majority of FCM experiments have been
analyzed manually by direct data inspection in one or two dimensions,
or by very basic comparisons of summary statistics. We believe that
these approaches do not fully address the highly complex nature of FCM
data, in particular, they discard many of the fundamental aspects of
the data, such as its underlying distribution and its high-dimensional
nature. In addition, the subjective character of manual analyses are a
major obstacle to reproducibility. For FC-HCS data, unassisted manual
inspection has become time and labor intensive, and robust
statistical methods need to be developed to point the investigator to
the interesting aspects of the data or to potential problems. While
the expert knowledge of immunologists and researchers remains crucial
for the understanding of FCM data, we believe that close collaboration
with other research fields such as statistics and computer science can
greatly improve the relevance of FCM in today's
high-throughput paradigm.

These issues are currently being addressed by emerging new flow
cytometry data standards developed in collaboration with the ISAC Data
Standards Task Force.  For example, the Gating-ML Candidate Recommendation (CR)
represents a proposal on how to form unambiguous XML-based gate
definitions, which can facilitate the interchange and validation of
data between different software packages with the potential of
significant increase of hardware and software interoperability. Gates
may be ordered into a hierarchical structure to describe a gating
strategy. They may be applied on parameters as in FCS files or on
transformed parameters as described by an explicit parameter
transformation. This enables the encoding of a simple analytical work
flow, in a way that allows us to reconstruct the analysis
programmatically.

The flowCore framework presented here offers import functionality for
raw data FCS files along with their complete set of file-specific meta
information (Figure~\ref{fig1:FrameWork}). Moreover, it is a software
implementation of the Gating-ML CR, which opens the possibility to
integrate flowCore in existing work flows and to communicate with
 every other FCM tool that adheres to the proposed
standard.

flowCore is not a GUI-driven software, and all operations are done
using a command line interface. Conceptually, it is possible to add a
more elaborate user interface on top of that, however the focus is on
a programmatic approach to enable the convenient development of novel
analysis methods. By taking the burden of data management from the
programmer, and by providing well-defined APIs and a structured class
hierarchy, it is possible to readily test new ideas and to easily
extend the frameworks functionality. %Both the underlying R and
%Bioconductor software and flowCore itself are open source and open %already mentioned its open source
%development projects, thus community involvement in the developing
%process is highly appreciated.

\section*{Basic data structures}
\subsection*{Flow Frame}
flowCore's primary task is the representation and basic manipulation
of FCM data. This is accomplished through a data model very
close to that adopted by other Bioconductor packages used for gene
expression analysis, and thus familiar to most Bioconductor
users. Similar to any other FCM data analysis tool, the basic unit of organization
in flowCore is a collection of events. We call the structure that hold
this data a flow frame. All relevant information for the collection
are stored along with the matrix of raw data values and can be
accessed programmatically. Most commonly, these are descriptors of the
stains used in the experiment and the respective measurement channels,
information about compensation performed at the instrument side and
any additional keywords the user deems to be important to annotate the
data. A number of quality checks are performed during the creation of
a flow frame to ensure data integrity. There is an extended interface
that allows for the standardized interaction with the flow data, thus
enabling the development of more specialized tools for specific tasks
of the data analysis process that all share a common Application Program Interface (API).

\subsection*{Flow Set}
In high-throughput FCM, many of the analysis tasks need to
be performed consistently across multiple flow measurements, hence we
introduce the concept of a collection of flow frames called a flow
set. Flow sets store the relevant information associated with each
individual frame such as descriptions of the samples or the treatment
to which a sample was subjected, and they manage the consistent
application of operations on the individual flow frames. The burden of
keeping score of the annotation is shifted from the user to the
infrastructure, thus reducing the risk of common errors that occur when using more or less ad hoc spreadsheets that lack a
controlled structure. In general, there is no native ordering to a
flow set, which in a way reflects the still predominant application of
tube-based experiments rather than experiments following a
plate-design. However, the flow set structure can be readily extended
to also accommodate such spatial information \textit{(plateCore
  reference, Perry, Errol?)}.

%%Figure1 Diagram of the data structures and the basic operations
\begin{figure}
\centering
\includegraphics[width=0.9\textwidth]{Figure1-flowCoreFrameWork.jpg}
\caption{\label{fig1:FrameWork}{The flowCore frame work. For each
    experiment, FCS files, phenotypic and meta data is stored in a
    flow set. Each flow frame of a flow set corresponds to one FCS
    file. All basic operations (e.g., compensation, transformation, gating) can be applied on both single flow frames or entire flow
    set.}}
\end{figure}

\section*{Standard flow operations}
The basic operations in all FCM analyses are typically the
same: the data need to be compensated (if not already done on the
instrument side) and transformed, and sup-populations of interest
need to be selected based on a set of (predominantly sequential)
gates. All software solutions for FCM analysis offer more or
less elaborate support for these kinds of operation, most of them in
an interactive, GUI-driven interface. The approach we have taken in
flowCore is to abstractly describe these operations and build a set of
tools to perform them on both flow frames as well as on entire flow
sets. While transformation, and to a certain extent compensation,
are fairly routine operations with only limited potential for
improvement, being able to implement new methodologies for gating of
FCM data, and extending the capabilities of flowCore
through object oriented programming are features that clearly sets our
framework apart from other FCM analysis tools. By factoring out as
much of the bookkeeping as possible, programmers can focus on the
actual operations rather then having to deal with the tedious details
of data integration and access. Third-party methods can act on their own as first-class citizens in the analysis
framework, without breaking the workflow or the basic
infrastructure. This design allows for the straightforward extension
of flowCore's capabilities, and has already fostered the development of a
number of valuable add-ons. \citep{lo2008agf,sarkar2008ufv}
\textit{(plateCore reference, Perry, Errol?)}.

\subsection*{Transformation and compensation}
Data transformation is essential for both data
visualization and modelling \citep{lo2008agf}. The major
transformations that are routinely used in FCM analysis
have been implemented in flowCore <e.g., subset list?>. Furthermore, the design of the R
language makes it easy to define arbitrary functions to apply to the
data of individual flow frames or entire flow sets,
respectively. Compensation is available for both flow frames and flow
sets, and the software also offers functionality to compute spillover
or compensation matrices from a set of appropriate compensation
samples.

@@

Static gating for all samples in a high-throughput FCM
experiment is often not an option, since the measured variables tend
to vary between different treatments or over time. Automated or
data-driven gating tries to estimate the gating regions from the
underlying data, thus providing a fast objective solution to the
analysis of potentially very large and diverse data sets
\cite{lo2008agf}. One of the automated gating methods implemented in
flowCore is based on identifying areas of significant curvature in a
kernel density estimate of the data \citep{wand2008}. Assuming that
the regions of interest are of high density, the software is able to
reliably detect them in either the one ore the two-dimensional density
landscape. Kernel density estimation is a well-known problem in
statistical computing, and a lot of effort has been invested in the
development of good software to address it. The modular design of R,
Bioconductor and flowCore allows to easily integrate these existing
solutions into our framework. In this example, we directly use R code
from the feature package developed by Wand et al. Instead of
re-writing existing code, we are able to include it via the well
tested distribution mechanism provided by R's software package
system. This process is bi-directional, and all functionality
implemented in flowCore is available to other package authors, as we
have seen with the afore-mentioned flowClust package.

Through flowCore, we have provided the FCM community a freely available, sophisticated, development and
analysis platform for high throughput data analysis. The scope and complexity of
the technology is evolving rapidly, and a collaborative effort in
devising new methodolgy has been proven beneficial for a number of
different biological and computational biology challenges. We hope
that our framework will be the foundation for fruitful shared research
by many calloborators from multiple scientific fields, further pushing
the boundaries of the exiting and still promising FCM
technology.

\section*{Acknowledgements}
This work was supported by NIH grant EB005034. RRB is a Michael Smith Foundation for Health Research Scholar and a International Society for Analytical Cytology Scholar.
